Supplementary materials

### 2021-09-16 manuscript.Rmd
sup mat & minor changes to paper

```{r fig-xling-freq-age1, fig.pos = "!t", fig.align = "center", out.width = '65%', fig.cap="(WE WILL PROBABLY NOT USE THIS ONE) Model predicting NWR scores from the average frequency with which each phone is found across languages (x axis) and child age (median split on scaled age)."}
plot_model(xl_fr,type="int")
```

```{r fig-xling-freq-age2, fig.pos = "!t", fig.align = "center", out.width = '65%', fig.cap="(WE WILL PROBABLY NOT USE THIS ONE) Model predicting NWR scores from the average frequency with which each phone is found across languages (x axis) and child age (mean, or plus/minus one standard deviation)."}

plot_model(xl_fr,type="pred",terms=c("avg_fr_s","age.rounded_s"))
```

>>>>

<!-- ac2mc if it's easy for you to add dots representing actual children in the above, that would be awesome! If not, I can try to tackle it :) -->

### 2021-09-16 supmat.Rmd
end r3
sup mat & minor changes to paper

---
title: "Supplementary materials"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TODO
We have included information about the stimulus length in the caption of the figures.

## SM1: Systematic review of previous work to check whether length there are typological differences in length effects on NWR

@@

ADD MORE INFO ABOUT HOW THESE PAPERS WERE DISCOVERED

 We do not think previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths, as described in Supplementary Materials. We found data for learners of Israeli Arabic  [@jabere2018xperiment]; Cantonese  [@stokes2006nonword]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant]. 
Our reading of that work is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically line up with expected word length in different languages. For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (~`r round(difs[difs$Group.1=="Tsimane'","x"])`%) and Arabic (~`r round(difs[difs$Group.1=="Arabic","x"])`%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only ~`r round(difs[difs$Group.1=="Italian","x"])`%. Similarly, two languages that are often described as heavily biased towards monosyllables show diverse length effects (Cantonese ~`r round(difs[difs$Group.1=="Cantonese","x"])`% versus English ~`r round(difs[difs$Group.1=="English","x"])`%).

@@

## SM2: Integrating data from other studies

To enrich this discussion, we looked for previous studies on monolingual children with normative development learning diverse languages, and entered them when they reported non-word repetition scores based on whole item scoring. We entered data from XX studies (including ours), presenting data from XX languages. Specifically, Arabic was represented by @jabere2018xperiment; Cantonese by @stokes2006nonword; English by @vance2005speech; Italian by @piazzalunga2019articulatory; Mandarin by @lei2011developmental; Persian by @farmani2018normalization; Slovak by @kapalkova2013non and @polivsenska2014improving; Sotho by @wilsenach2013phonological; Spanish by @balladares2016socio; Swedish by @kalnak2014nonword and @radeborg2006swedish; Tsimane' by @cristia2020infant; and Yélî Dnye from the present study. Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability. Results separating different age groups are shown in Figure \@ref(fig:fig-prevlit-overall).

FIGURE REMOVED

@@

## SM2: Age effects on accuracy

```{r fig-xage, fig.pos = "!t", fig.align = "center", out.width = '65%', fig.cap="NWR scores for individual target words as a function of the child's age (averaged in years)."}
aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$target,round(trnsc.inc$age.rounded)),mean)->xage
colnames(xage)<-c("Target","Age","Word")
#stims.ipa <- read.xlsx("stimuli.xlsx") %>%
#  select(target, target_ortho, phono)
xage <- merge(xage, stims.ipa, by.x = "Target", by.y = "target")

item.plot <- ggplot(data = xage, aes(x = Age, y = Word, label = phono)) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  geom_text(family = "Doulos SIL") +
  ylab("NWR scores (% words)") +
  xlab("Children's age") +
  theme_apa()
ggsave("nwr.by.age.ITEM.png", dpi = 300, device = png())

knitr::include_graphics("nwr.by.age.ITEM.png")

>>>>

---
title             : "Supplementary materials to: Non-word repetition in children learning Yélî Dnye"
shorttitle        : "SM: NWR in Yélî Dnye learners"

author: 
  - name          : "Alejandrina Cristia"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "29, rue d’Ulm, 75005 Paris, France"
    email         : "alecristia@gmail.com"
  - name          : "Marisa Casillas"
    affiliation   : "2,3"
    address       : "Wundtlaan 1, 6525 XD, Nijmegen, The Netherlands"
    email         : "marisa.casillas@mpi.nl"

affiliation:
  - id            : "1"
    institution   : "Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes Cognitives, ENS, EHESS, CNRS, PSL University"
  - id            : "2"
    institution   : "Max Planck Institute for Psycholinguistics"
  - id            : "3"
    institution   : "University of Chicago"

author_note: >
       Both authors contributed to study funding, design, data collection, annotation, analyses, writing.

abstract: >
   This is a supplementary material containing: 1. Explanation of a small-scale systematic review carried out to embed the main results in the NWR literature; 2. The use of this data to look at potential length effects on NWR as a function of language characteristics; 3. Length and age effects in this and previous work.

bibliography      : ["nwrbib.bib"]

figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no

lang              : "english"
class             : "doc"
output            :
  papaja::apa6_pdf
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(dplyr)
library(openxlsx)
```

## SM1: Small systematic review of previous NWR work

For many of our research questions, we wanted to know about previous work. It is common in psychology to do this via an unsystematic review -- that is, by e.g. starting out with papers one knows and then reading other papers citing those, or by doing searches on google scholar or another search engine, or by reading someone else's review. It has been well-established that overviews based on such unsystematic searches do not generalize and have other flaws because they represent biased samples of the literature [@thomas2020impact] -- for instance, famous studies are more likely to be found, even though there are no quality differences between more versus less famous studies.

So instead of relying on such an approach, we built on a systematic review that was carried out for a different paper: @cristia2020infant "combined our previous knowledge of the literature and systematic searching to yield a sample of 17 studies that we could interrogate further". Specifically, the main goal of that systematic review was to check whether "effects of infant-directed input quantities may have been reported. Input variability has not been studied, to our knowledge, but two potential proxies of input have: Monolingual status, and socio-economic status.  We set aside the literature comparing monolinguals against non-monolinguals because we reasoned that some readers may argue this contrast does not only show effects of input differences, but also interference across the languages being learned" (Cristia, 2020, personal records). Discovery combined an initial list produced by Gianmatteo Farabolini (based on readings of literature on bilingual NWR mainly), with a scholar.google.com search carried out in incognito mode by Cristia, with keywords "non-word-repetition socio-economic-status". Studies finally included reported on children who were monolingual and between 3 and 7 years of age, and which used proportion of non-words repeated as the outcome metric.

We revisited this selection of articles in order to extract data on actual performance, taking into account also the length of the non-words used in the study. From that selection, only 8 articles reported on actual performance in the text, figures, or tables, fortunately representing a range of languages: Persian by @farmani2018normalization; Israeli Arabic  [@jabere2018xperiment];  English  [@vance2005speech]; Slovak by @kapalkova2013non and @polivsenska2014improving; Sotho by @wilsenach2013phonological; Swedish by @kalnak2014nonword and @radeborg2006swedish. We extended that selection with 5 additional papers selected to increase the number of languages represented, with both languages sometimes described as having short words [Cantonese  @stokes2006nonword; Mandarin  @lei2011developmental]; and others described as having long words [Italian  @piazzalunga2019articulatory; Spanish @balladares2016socio]. Finally, we also included Tsimane' data  [@cristia2020infant].

## SM2: Assessing potential length effects

```{r demo}
# please note that you will not be able to do this because you don't have this version, that is not fully de-identified
# read.csv("NWR-test-log_FINAL.csv",header=T,sep=";")->fulldemo
# fulldemo=fulldemo[,-2] #remove children's names
# fulldemo$dob=as.Date(as.character(fulldemo$DOB.ISO),format="%Y%m%d")
# fulldemo$test=as.Date(as.character(fulldemo$Date),format="%Y%m%d")
# fulldemo$age.rounded=as.numeric(difftime(fulldemo$test,fulldemo$dob,unit="weeks"))/52.25
# fulldemo$Age=as.numeric(as.character(fulldemo$Age))
# plot(fulldemo$Age~fulldemo$age.rounded)
# write.csv(fulldemo,"nwr-demo.csv")

# read.csv("nwr-demo.csv",header=T)->demo
# demo=demo[,-c(20,21)] #remove DOBs 
# write.csv(demo,"nwr-demo.csv")

read.csv("nwr-demo.csv",header=T)->demo
#colnames(demo)

subset(demo,included=="yes" & Age <12 & !is.na(Age))->inc
#if age.rounded is missing, use reported age
inc$age.rounded[is.na(inc$age.rounded)]<-inc$Age[is.na(inc$age.rounded)]
#summary(inc)
table(inc$birthOrder)->birthorder.tab

```{r readin}
read.table("final_data.txt",header=T,encoding="UTF-8")->trnsc

trnsc$Sex=factor(trnsc$Sex)
trnsc$birthOrder=as.numeric(trnsc$birthOrder)

# Include only included children in trnsc
trnsc[which(trnsc$id %in% inc$ID),]->trnsc.inc

```{r exclude subsequent}
trnsc.inc-> trnsc.inc_all #keep a copy for figure comparing first & subsequent

trnsc.inc=trnsc.inc[trnsc.inc$attempt=="first",]

@@

We looked at the subset of papers that reported NWR scores separately for different word lengths. These were: Israeli Arabic  [@jabere2018xperiment]; Cantonese  [@stokes2006nonword]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant].

Our reading of that work is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically line up with expected word length in different languages. For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (~`r round(difs[difs$Group.1=="Tsimane'","x"])`%) and Arabic (~`r round(difs[difs$Group.1=="Arabic","x"])`%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only ~`r round(difs[difs$Group.1=="Italian","x"])`%. Similarly, two languages that are often described as heavily biased towards monosyllables show diverse length effects (Cantonese ~`r round(difs[difs$Group.1=="Cantonese","x"])`% versus English ~`r round(difs[difs$Group.1=="English","x"])`%).

@@

## SM3: Integrating our data with that from other studies

For this analysis, we could include all studies that reported non-word repetition scores based on whole item scoring for at least some length, ideally separating children by age. Specifically, Arabic was represented by @jabere2018xperiment; Cantonese by @stokes2006nonword; English by @vance2005speech; Italian by @piazzalunga2019articulatory; Mandarin by @lei2011developmental; Spanish by @balladares2016socio;  Tsimane' by @cristia2020infant; and Yélî Dnye from the present study. Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability. Results separating different age groups are shown in Figure \@ref(fig:fig-prevlit-overall).

@@

# References

\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}
